---
title: "Reproducible Research Assig 1"
author: "Karin Kolbe"
date: "08/Nov/2020"
output: 
        html_document: 
                keep_md: true

---


## Loading and preprocessing the data
```{r}
activity <- read.csv("activity.csv",stringsAsFactors=FALSE)
```

Quick check of the data
```{r}
str(activity)
```

```{r}
summary(activity)
```


Load Packages
```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
```

## 1. What is mean total number of steps taken per day?

Group by date
```{r}
activity1 <- activity %>%
        group_by(date) %>%
        summarise(daily = sum(steps, na.rm=TRUE))
```
Check the data
```{r}
head(activity1)
```
Create a histogram of the total number of steps taken per day
```{r}
ggplot(activity1, aes(x = daily)) +
    geom_histogram(fill = "blue",binwidth=1000) +
    labs(title = "Daily Steps", x = "Steps", y = "Frequency")
```

```{r}
theMean   <- mean(activity1$daily,na.rm=TRUE)
theMean
theMedian <- median(activity1$daily,na.rm=TRUE)
theMedian
```

The mean of the steps is `r theMean`.   
The median of the steps is `r theMedian`  


## 2. What is the average daily activity pattern?

Group the data by 5-minute intervals
```{r}
activity2 <- activity %>%
        group_by(interval) %>%
        summarise(meanSteps = mean(steps,na.rm=TRUE))
```

Check what I have
```{r}
head(activity2)
```


```{r}
with(activity2, plot(interval,meanSteps, type="l",lty=1,lwd=2,col="blue",
     ylab="Average Steps", xlab="Interval",
     main="Time Series of the 5-minute Intervals"))

```

```{r}
intervalwithMax <- activity2  %>% 
       filter(meanSteps == max(activity2$meanSteps)) %>% 
        pull(interval)
intervalwithMax
```

The interval with maximum average number of the steps is `r intervalwithMax`.  


## 3. Imputing missing values
Detemine how many rows have NA
```{r, results="hide"}
totNAs <-sum(is.na(activity$steps))
```
There are `r totNAs` rows with missing values.

Replace the missing values with the average for the interval. This was calculated 
in step 2 and is in dataframe activity2.

```{r}
activity3 <- activity %>%
        left_join(activity2) %>%
        mutate(newSteps = ifelse (is.na(steps), meanSteps, steps))
head(activity3)
```

Group and total the data, and plot
```{r}
activity4 <- activity3 %>%
        group_by(date) %>%
        summarise(daily = sum(newSteps))

ggplot(activity4, aes(x = daily)) +
    geom_histogram(fill = "red",binwidth=2000) +
    labs(title = "Daily Steps", x = "Interval", y = "Frequency")
```


```{r}
theMean2   <- mean(activity4$daily)
theMean2
theMedian2 <- median(activity4$daily)
theMedian2
```
  
###Summarise the Mean and Median results
  
Version | Mean  | Median
--- | --- | ---
With NAs | `r theMean` | `r theMedian`
NAs replaced by interval mean| `r theMean2` | `r theMedian2`
  
Both the mean and median have increased slightly by replacing the NAs, which is 
expected.
  
## Are there differences in activity patterns between weekdays and weekends?
```{r}

activity5 <- activity3 %>%
        mutate(dayType = 
                as.factor
                (ifelse ((wday(as.Date(date, format = "%Y-%m-%d")) %in% c(7,1)), 
                        "weekend", 
                        "weekday")))
head(activity5)

```


Group the data by 5-minute intervals and plot
```{r}
activity6 <- activity5 %>%
        group_by(dayType,interval) %>%
        summarise(meanSteps = mean(newSteps))

ggplot(activity6, aes(interval,meanSteps,color=dayType)) + 
        geom_line() + 
        labs(title = "Average Steps per Interval by Type of Day", 
         x = "Interval", y = "Average Steps") + 
        facet_wrap(dayType~., ncol = 1, nrow=2)
```

From this plot we can see that people take more steps earlier on weekdays, but 
then are more stationary during the day. 
On weekends people presumably stay long in bed, but are more active during the
day. They seem to go to bed a little later on weekends than weekdays on this plot.